---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

# explicaR

> *explicar* (Spanish) — to explain

Auto-document any R or Python project. Point `explicar()` at a codebase and get a browsable wiki, a dependency graph, and a searchable code index — with no manual writing.

```r
install.packages("explicaR")  # CRAN
# or: remotes::install_github("CathalByrneGit/explicaR")

explicar("path/to/any/project")   # opens in browser
```

---

## What you get

| Output | What it is |
|---|---|
| **Wiki** | One LLM-generated page per source file — overview, key functions, usage |
| **Graph** | Interactive Mermaid dependency diagram across scripts, functions, variables |
| **Index** | DuckDB code graph, queryable from Claude Desktop via MCP or in-browser SQL |
| **llms.txt** | Machine-readable file listing so any LLM understands the project instantly |

Everything runs locally. No code leaves your machine unless you choose a cloud LLM provider.

---

## Quick start

```{r quickstart}
library(explicaR)

# Static HTML viewer — no server, works offline
explicar("path/to/project")

# With LLM wiki pages (local Ollama — free, private)
explicar("path/to/project", llm = TRUE)

# Bring your own LLM
library(ellmer)
explicar("path/to/project", llm_chat = chat_anthropic())

# Remote repo
explicar("https://github.com/tidyverse/dplyr")
```

Any [ellmer](https://ellmer.tidyverse.org) provider works: Ollama, OpenAI, Anthropic, Gemini, Groq, AWS Bedrock, or any OpenAI-compatible endpoint.

---

## Claude Desktop / Claude Code (MCP)

Make your codebase queryable from Claude without copying and pasting code:

```{r mcp}
# Build the index once
explicar_index_build("path/to/project")
```

Add to `claude_desktop_config.json`:

```json
{
  "mcpServers": {
    "my-project": {
      "command": "Rscript",
      "args": ["-e", "explicaR::serve_explicar_mcp('path/to/project')"]
    }
  }
}
```

Claude can then search your code, run SQL on the graph, read wiki pages, and list files — without you explaining anything.

---

## Build the wiki and search it

```{r wiki-and-search}
library(ellmer)

# Generate wiki pages (stored in .explicar/explicar.duckdb)
explicar_wiki_build("path/to/project",
  llm_chat = chat_ollama(model = "llama3.2"))

# Ingest wiki + docs into the RAG store
explicar_ingest("path/to/project")

# Search — BM25 keyword or vector similarity (with Ollama embeddings)
explicar_doc_retrieve("how does parse dispatch work")
explicar_index_retrieve("authentication functions", top_k = 5)
```

---

## Viewer tiers

| Tier | What | How |
|---|---|---|
| **1** | Self-contained HTML — wiki, graph, verb tables | `explicar()` |
| **2** | Same + in-browser SQL via DuckDB-WASM | `generate_wasm_viewer()` |
| **3** | Live server with streaming chat (SSE) | `view_explicar_db()` |

---

## Privacy

`explicar()` and `explicar_parse()` make no network calls. The browser
loads Mermaid from CDN. LLM calls only happen if you pass `llm = TRUE`
or `llm_chat = ...`.

---

## License

MIT
