---
title: "Getting Started with explicaR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with explicaR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

explicaR is an **R DeepWiki**: point it at any local R or Python project (or a
GitHub URL) and get a self-contained, browsable, searchable wiki — one page per
source file, an interactive Mermaid dependency graph, a DuckDB code index, and
an optional live chat server backed by a ragnar RAG store.

```{r install, eval=FALSE}
# Install from GitHub
remotes::install_github("CathalByrneGit/explicaR")
```

## One-line quick start

```{r quickstart}
library(explicaR)

# Analyse a local project and open the viewer in the browser
explicar("path/to/your/project")
```

This single call:

1. Parses every `.R` (and optionally `.py`) file into a code graph
2. Generates an interactive Mermaid dependency diagram
3. Writes a self-contained `explicar_viewer.html` and opens it

No LLM is required for this step — everything runs offline.

## Parsing and the graph

The two core primitives are available independently:

```{r parse}
# Parse the project into a code graph
pr <- explicar_parse("path/to/project")

# What you get back:
pr$nodes   # tibble: name, type, file, line, label, shape_info
pr$edges   # tibble: from, to, type
pr$verbs   # tibble: file, line, fn_name, input_var, output_var, pkg

# Get the Mermaid diagram text (paste into GitHub, Obsidian, Quarto…)
cat(explicar_graph(pr))
cat(explicar_graph(pr, direction = "LR"))
```

### Node types

| Shape | Colour | Type | Meaning |
|-------|--------|------|---------|
| Rectangle `[…]` | Blue | `script` | `.R` / `.py` source file |
| Rounded `(…)` | Green | `variable` | Named object (LHS of assignment) |
| Hexagon `{{…}}` | Orange | `function` | Named function definition |
| Cylinder `[(…)]` | Purple | `source` | Data file or Python import |

### Edge types

`produces` · `consumes` · `calls` · `reads` · `writes` · `depends`

## Three viewer tiers

### Tier 1 — Self-contained HTML (default)

Zero runtime dependencies.  Works offline; only the Mermaid CDN request
requires internet.

```{r tier1}
generate_viewer(pr, output_file = "my_wiki.html")
```

### Tier 2 — WASM SQL browser

Embeds DuckDB-WASM so you can run SQL against the code graph directly in the
browser — no server needed.

```{r tier2}
generate_wasm_viewer(pr, output_file = "my_wiki_wasm.html")
```

Example queries you can run in the browser:

```sql
-- All function definitions
SELECT name, file, line FROM nodes WHERE type = 'function' ORDER BY name

-- What calls what
SELECT from_node, to_node FROM edges WHERE type = 'calls'

-- Pipeline verbs in file order
SELECT fn_name, file, line FROM verbs ORDER BY file, line
```

### Tier 3 — Live server with chat

Starts an HTTP server at `http://127.0.0.1:8080` with a full analytics
dashboard: the Mermaid graph, wiki pages, keyword search, and an LLM chat
panel.

```{r tier3}
library(ellmer)

view_explicar_db(
  project_dir = "path/to/project",
  llm_chat    = chat_ollama(model = "llama3.2"),  # any ellmer provider
  port        = 8080
)
# → opens http://127.0.0.1:8080
```

## Mixed R and Python projects

```{r python}
pr <- explicar_parse("path/to/project", languages = c("r", "python"))
explicar("path/to/project", languages = c("r", "python"))
```

## Remote repositories

```{r remote}
# Clone from GitHub, parse, and open the viewer
explicar("https://github.com/tidyverse/dplyr")

# Private repos — set GITHUB_PAT or pass explicitly
Sys.setenv(GITHUB_PAT = "ghp_...")
explicar("https://github.com/myorg/private-repo")

# Use the cached clone without pulling again
explicar("https://github.com/tidyverse/dplyr", update = FALSE)
```

Clones land in `~/.explicar/repos/<host>/<owner>/<repo>/`.  Subsequent calls
run `git pull` unless `update = FALSE`.

## Building the persistent code index

For search and MCP integration you need the DuckDB index:

```{r index}
# Build (incremental — only re-parses changed files)
explicar_index_build("path/to/project")

# Keyword search
explicar_index_retrieve("functions that filter rows", top_k = 5)
```

The index lands in `.explicar/explicar.duckdb` inside the project directory.

## What's next?

- **LLM wiki pages and deep research** — see `vignette("llm-wiki")`
- **MCP integration with Claude Desktop** — see `vignette("mcp-claude")`
