---
title: "LLM Wiki Pages and Deep Research"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LLM Wiki Pages and Deep Research}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

explicaR can generate a markdown wiki page for every source file in your
project using any LLM supported by the
[ellmer](https://ellmer.tidyverse.org) package.  Pages are stored in a DuckDB
table and ingested into a ragnar RAG store for keyword and vector search.

## LLM provider support

All wiki functions accept any `ellmer::Chat` object.  Pick whichever provider
suits your workflow:

| Provider | Call | Notes |
|---|---|---|
| **Ollama** (local) | `chat_ollama("llama3.2")` | Free, private, no key |
| **OpenAI** | `chat_openai(model = "gpt-4o-mini")` | Reads `OPENAI_API_KEY` |
| **Anthropic** | `chat_anthropic(model = "claude-haiku-4-5")` | Reads `ANTHROPIC_API_KEY` |
| **Google Gemini** | `chat_google_gemini()` | Reads `GOOGLE_API_KEY` |
| **Groq** | `chat_groq()` | Reads `GROQ_API_KEY` |
| **AWS Bedrock** | `chat_aws_bedrock()` | Uses AWS credentials |
| **llama.cpp / vLLM** | `chat_openai_compatible(base_url = "http://...")` | Any OpenAI-compatible endpoint |

## Generating wiki pages

```{r wiki-build}
library(explicaR)
library(ellmer)

# Ollama — completely local, no cost
explicar_wiki_build(
  "path/to/project",
  llm_chat = chat_ollama(model = "llama3.2")
)

# OpenAI
explicar_wiki_build(
  "path/to/project",
  llm_chat = chat_openai(model = "gpt-4o-mini")
)

# Anthropic
explicar_wiki_build(
  "path/to/project",
  llm_chat = chat_anthropic(model = "claude-haiku-4-5")
)
```

Each page contains four sections:

1. **Overview** — 2–3 sentences on what the file does
2. **Key Functions / Objects** — bullet list with one-line descriptions
3. **How It Works** — implementation approach and key patterns
4. **Usage Example** — short code block showing typical use

A per-file Mermaid dependency diagram is prepended when the file has
connections in the graph.

### No LLM available?

Pass `llm_chat = FALSE` for the roxygen fallback — pages are built from
`#'` comments and the parse graph, with no network calls:

```{r fallback}
explicar_wiki_build("path/to/project", llm_chat = FALSE)
```

## Change detection

`explicar_wiki_build()` compares each file's `mtime` against the value stored
in the `wiki` table.  Unchanged files are skipped automatically:

```{r change-detection}
# First run — generates all pages
explicar_wiki_build("path/to/project", llm_chat = chat_ollama("llama3.2"))

# Second run — skips files whose mtime hasn't changed
explicar_wiki_build("path/to/project", llm_chat = chat_ollama("llama3.2"))
#> skip (unchanged): foo.R
#> skip (unchanged): bar.R
#> Wiki complete: 0 generated, 2 unchanged

# Force regeneration of everything
explicar_wiki_build("path/to/project",
                    llm_chat = chat_ollama("llama3.2"),
                    force    = TRUE)
```

## Where pages are stored

Wiki pages live in the `wiki` table of `.explicar/explicar.duckdb`:

```sql
SELECT file, model, generated_at, length(content) AS chars
FROM wiki
ORDER BY file
```

You can query this directly with DBI / duckdb:

```{r query-wiki}
library(DBI)
library(duckdb)

con <- dbConnect(duckdb(), ".explicar/explicar.duckdb", read_only = TRUE)
wiki <- dbGetQuery(con, "SELECT file, model, content FROM wiki")
dbDisconnect(con, shutdown = TRUE)

cat(wiki$content[[1]])  # print the first page
```

## Ingesting into the RAG store

After building wiki pages, ingest them (plus README, roxygen docs, vignettes)
into the ragnar BM25 store so they are searchable:

```{r ingest}
explicar_ingest("path/to/project")
```

Then search with keyword matching:

```{r search}
explicar_index_retrieve("functions that filter rows", top_k = 5)
```

## Vector embeddings (semantic search)

`explicar_embed()` converts every code-graph node to a natural-language
description and embeds it in the ragnar VSS store using Ollama.  This enables
semantic search — finding relevant code even when exact keywords don't match.

```{r embed}
# Requires Ollama with an embedding model
explicar_embed("path/to/project", model = "nomic-embed-text")

# Semantic search — finds conceptually similar code
explicar_semantic_retrieve("function that removes duplicate rows")
explicar_semantic_retrieve("data cleaning pipeline")

# Force BM25 keyword search instead of VSS
explicar_semantic_retrieve("filter_rows", bm25_only = TRUE)
```

The node text format is:
```
function: `clean_data()` — Removes NA rows and duplicates [R/clean.R:14]
```

## Deep research

`deep_research()` runs a multi-turn investigation loop using an ellmer chat
with a ragnar retrieval tool registered on it.  The LLM plans its research,
iterates (retrieving relevant chunks on each turn), and terminates when it
writes a response starting with `CONCLUSION:`.

```{r deep-research}
library(ellmer)
library(ragnar)

# Connect to the store built by explicar_ingest() / explicar_embed()
store <- ragnar_store_connect(".explicar/explicar.duckdb")

# Build the chat with retrieval wired in
chat <- chat_ollama(model = "llama3.2")
ragnar_register_tool_retrieve(chat, store)

# Run up to 5 research iterations
deep_research(chat, "How does the parse dispatch layer work?",
              max_iterations = 5L)

# Print the final conclusion
cat(chat$last_turn()$text)
```

The loop:
1. **Plan turn** — LLM lists what it needs to investigate
2. **Iterate** (up to `max_iterations - 1` turns) — retrieves chunks,
   synthesises findings
3. **Terminates** when the response begins with `CONCLUSION:`

## Full pipeline

```{r full-pipeline}
library(explicaR)
library(ellmer)

chat <- chat_ollama(model = "llama3.2")

# 1. Build the code graph index
explicar_index_build("path/to/project")

# 2. Generate wiki pages (change-detection: skips unchanged files)
explicar_wiki_build("path/to/project", llm_chat = chat)

# 3. Ingest into the ragnar BM25 store
explicar_ingest("path/to/project")

# 4. Embed code-graph nodes for semantic search (requires Ollama)
explicar_embed("path/to/project")

# 5. Open the Tier 1 viewer with wiki pages included
explicar("path/to/project", llm_chat = chat, open = TRUE)

# 6. Or launch the Tier 3 live server with chat
view_explicar_db("path/to/project", llm_chat = chat)
```

## Generating llms.txt

`explicar_llms_txt()` produces a machine-readable `llms.txt` — a concise
project summary (stats, file list, function index, Mermaid DAG) that any LLM
can ingest in one shot:

```{r llms-txt}
explicar_llms_txt("path/to/project")
# → writes llms.txt in the project root
```
