---
title: "MCP Integration with Claude Desktop and Claude Code"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MCP Integration with Claude Desktop and Claude Code}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

explicaR includes a stdio [Model Context Protocol](https://modelcontextprotocol.io)
(MCP) server that wires your project's code graph and wiki pages directly into
Claude Desktop or Claude Code.  Once connected, Claude can answer questions
about your codebase without you copying and pasting code.

## How it works

`serve_explicar_mcp()` reads the `.explicar/explicar.duckdb` built by
`explicar_index_build()` and exposes four tools over JSON-RPC 2.0 stdio:

| Tool | What it does |
|---|---|
| `search_code` | Hybrid BM25 / keyword search over nodes, wiki pages, and docs |
| `query_graph` | Run `SELECT` SQL against `nodes`, `edges`, `verbs`, `functions`, `wiki` |
| `get_wiki` | Retrieve the LLM-generated wiki page for a specific file |
| `list_files` | List all source files with type counts and function summaries |

## Step 1: build the index

```{r build}
library(explicaR)

# Parse the project and build the DuckDB index
explicar_index_build("path/to/project")

# Optionally add wiki pages (richer results for get_wiki / search_code)
library(ellmer)
explicar_wiki_build("path/to/project", llm_chat = chat_ollama("llama3.2"))
explicar_ingest("path/to/project")
```

## Step 2: start the MCP server

```{r serve}
serve_explicar_mcp("path/to/project")
# Blocks on stdin — leave this running while Claude is connected.
```

## Step 3: add to claude_desktop_config.json

Add the following to your Claude Desktop configuration file.

**macOS**: `~/Library/Application Support/Claude/claude_desktop_config.json`
**Windows**: `%APPDATA%\Claude\claude_desktop_config.json`

```json
{
  "mcpServers": {
    "my-project": {
      "command": "Rscript",
      "args": ["-e", "explicaR::serve_explicar_mcp('path/to/project')"]
    }
  }
}
```

Restart Claude Desktop after editing the config.  The server name
(`"my-project"`) will appear in Claude's tool list.

### Multiple projects

Register each project under a distinct key:

```json
{
  "mcpServers": {
    "dplyr": {
      "command": "Rscript",
      "args": ["-e", "explicaR::serve_explicar_mcp('~/projects/dplyr')"]
    },
    "my-app": {
      "command": "Rscript",
      "args": ["-e", "explicaR::serve_explicar_mcp('~/projects/my-app')"]
    }
  }
}
```

## Using the tools from Claude

Once connected, Claude can use the tools transparently when you ask questions.
Example prompts that trigger tool use:

- *"What functions are defined in `clean.R`?"* → `get_wiki` or `search_code`
- *"Which files call `filter_rows()`?"* → `query_graph`
- *"How does the parse dispatch layer work?"* → `search_code` + `get_wiki`
- *"List all exported functions and their files."* → `query_graph`

### query_graph: SQL examples

Claude can run any read-only `SELECT` query.  Non-`SELECT` statements are
rejected.  The queryable tables are:

```sql
-- All function definitions
SELECT name, file, line FROM nodes WHERE type = 'function' ORDER BY name

-- Call graph: who calls what
SELECT from_node, to_node FROM edges WHERE type = 'calls'

-- Files that produce a specific variable
SELECT from_node, to_node FROM edges
WHERE type = 'produces' AND to_node = 'clean_df'

-- Functions with descriptions (populated by explicar_index_build)
SELECT name, file, line, description FROM functions

-- All wiki pages and their sizes
SELECT file, model, length(content) AS chars FROM wiki ORDER BY chars DESC
```

## Claude Code integration

For Claude Code (the CLI), add the server via the MCP settings:

```bash
# Add to project-level MCP config
claude mcp add my-project -- Rscript -e "explicaR::serve_explicar_mcp('.')"
```

Or edit `.claude/settings.json` manually:

```json
{
  "mcpServers": {
    "my-project": {
      "command": "Rscript",
      "args": ["-e", "explicaR::serve_explicar_mcp('.')"]
    }
  }
}
```

## Tier 3 live server: in-browser chat

For an interactive browser experience instead of the CLI integration, use
the Tier 3 server.  It provides the same search and chat capabilities in a
dashboard UI:

```{r tier3}
library(ellmer)
library(explicaR)

view_explicar_db(
  project_dir = "path/to/project",
  llm_chat    = chat_ollama(model = "llama3.2"),
  port        = 8080
)
# → http://127.0.0.1:8080
```

The dashboard includes:

- **Graph tab** — interactive Mermaid dependency diagram with clickable nodes
- **Wiki tab** — LLM-generated pages per file
- **Search tab** — BM25 keyword search over the ragnar store
- **Chat tab** — RAG-powered Q&A (requires `llm_chat`)

HTTP endpoints exposed by the server:

| Method | Path | Description |
|---|---|---|
| `GET` | `/` | Dashboard HTML |
| `POST` | `/chat` | Single-turn chat |
| `GET` | `/chat/stream` | Streaming chat (SSE) |
| `POST` | `/search` | Keyword search |
| `GET` | `/graph.json` | Nodes + edges as JSON |
| `GET` | `/wiki.json` | All wiki pages as JSON |
| `GET` | `/health` | Health check |

## Privacy

| Operation | Network calls |
|---|---|
| `explicar_index_build()` | None |
| `serve_explicar_mcp()` | None — reads local DB only |
| `view_explicar_db()` — no `llm_chat` | None |
| `view_explicar_db(llm_chat = chat_ollama(...))` | Local Ollama only |
| `view_explicar_db(llm_chat = chat_openai(...))` | OpenAI API |

The MCP server itself never makes network calls.  It only reads from
`.explicar/explicar.duckdb`.
